A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-10 latex/main.tex

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues}
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues.
Figure~\ref{fig:recombinant_qc} shows
the net number of supporting loci for the left and right parents
for all 855 recombination events.
The 501 recombinants having fewer than 4 net loci supporting each parent
are highly enriched for samples sequenced using the AmpliSeq V1 protocol
% https://github.com/jeromekelleher/sc2ts-paper/issues/808
(70\% of the recombinants despite making up 2\% of the Viridian dataset), and
71 (14\%) are associated with 6-base deletion in the Delta lineage,
known to be affected by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and
the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).
This QC filter also correlates strongly with results of other 
recombination detection methods when run on the putative 
recombinant sequences (see next section).

\subsection*{Validating and evaluating plausibility of recombination events}
[THIS IS VERY MUCH WIP]

We validated the recombination events in the sc2ts ARG by running 
several recombination detection programs on the putative recombinant
sequences. [TODO generate supplemental table summarising results of 
valiation.]

We first ran 3SEQ~\cite{Lam2018improved} [etc etc].

To assess the evidence for these putative novel recombination events we 
ran CovRecomb~\cite{Li2024CovRecomb} and rebar on the corresponding sequences 
(STAR Methods). Of the 354 QC-passing sc2ts recombination events, 
90 are classified as 
recombinant by CovRecomb (all but 2 of the 510 QC-failing recombinants 
were classified as non-recombinant).
Rebar classified 203 of the QC-passing sc2ts recombination events, 
as recombinant, with only 10 of those failing QC being classified 
as recombinant.

>>>>

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions \cite{DeMaio2020issues}
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
We developed a quality control (QC) measure
which counts the number of ``loci'' (clusters of closely spaced sites)
supporting the left and right parents of a recombinant (STAR Methods),
and effectively identifies a range of issues.
Figure~\ref{fig:recombinant_qc} shows
the net number of supporting loci for the left and right parents
for all 855 recombination events.
The 501 recombinants having fewer than 4 net loci supporting each parent
are highly enriched for samples sequenced using the AmpliSeq V1 protocol
% https://github.com/jeromekelleher/sc2ts-paper/issues/808
(70\% of the recombinants despite making up 2\% of the Viridian dataset), and
71 (14\%) are associated with 6-base deletion in the Delta lineage,
known to be affected by amplicon dropout\cite{Sanderson2021}.
The 354 remaining recombination events that pass this QC filter
contain all of the recombinants associated with Pango X lineages
(Table~\ref{tab:pango_x_lineages}) and
the Jackson et al. recombinants (Document~\ref{sec:jackson_recombs}).

\subsection*{Quantification of evidence for recombination}
